Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-05-10 paper.tex
Attempted an abstract. Removed the rambling about complexity. The extra rendered space is brutal.

\begin{abstract}
  Lorem Ipsum
\end{abstract}

@@

Once we finitize the address space of the store, this computes a
characteristic finite state machine that has only reachable states of
the pushdown system the unbounded stack model embodies.
To turn this semantics into PDCFA's (implemented) algorithm for constructing a Dyck
state graph, we apply a widening operator ${\mathcal F}$ to make $\mstore$, $\mkont$,
and $\mmemo$ shared amongst all states.

@@

\paragraph{Example:} Let's consider a monovariant allocation strategy
(the address for each binding is the variable name itself) to run the
following example:
\begin{lstlisting}[mathescape]
  (let* ([id ($\lambda$ (x) x)]
         [y (id 0)]
         [z (id 1)])
    ($\le$ y z))
\end{lstlisting}
We should expect that a pushdown analysis would predict this evaluates
to true, and there are no loops in the program. 0CFA claims there is a
loop from the second call of \texttt{id} to the first, and thusly
predicts this program evaluates to true or false. PDCFA claims there
is no loop, and depending on the implementation, that the result is
true or false (paper), or just true (implemented). Let's take a look
at the evaluation after binding \texttt{id} ($\mstore_0 = [\texttt{id} \mapsto \set{\vclo{\slam{\texttt{x}}{\texttt{x}}}{\bot}}]$):

@@

\paragraph{PDCFA paper versus implementation:} since PDCFA is a
simpler analysis than CFA2, it enjoys a polynomial complexity in the
monovariant case. Its final ${\mathcal O}(n^6)$ form is a naive ``step
all states'' widening, which yields poor results as evident in the
above example. The common widening uses a work-list to only step
states that need stepping like the frontier-based widening we use ---
the fact that the store is shared amongst all states makes this a
semantic difference, and not just an implementation difference. Doing
this bumps the complexity up to ${\mathcal O}(n^9)$. The available
\emph{implementations} of PDCFA indeed do this and are themselves
${\mathcal O}(n^9)$. These bounds are (rough) absolute worst case that
is highly unlikely, if ever, to happen in practice. The work-list
abstraction restricts value flow to fewer spurious paths, often
leading to a faster analysis than a typical ${\mathcal O}(n^3)$ 0CFA
~\citep{ianjohnson:vardoulakis-lmcs11,
  dvanhorn:Earl2012Introspective}.

@@

CFA2 uses a ``local semantics'' that is similar to our segmenting
continuations at function boundaries, but gets stuck when it gets to a
point where it would need to ``return.'' It instead appeals to an
external, imperative algorithm for summarization to sew the function
calls and returns together. What we show here is almost wholly the
same, only embodied still in terms of a machine semantics and thus
more easily reasoned about. The analogy to memoization as
summarization guide us in the construction of the analysis in the
coming section.

@@

The way we have been splitting continuations at function calls has
similarities to the meta-continuation approach to modeling delimited
control, given in figure \ref{fig:shift-reset} (adapted from
~\citep{ianjohnson:Biernacki2006274}). We could view each function
call as inserting a prompt, and returns as aborting to the nearest
prompt. The changed semantic spaces for the shift/reset semantics are as
follows:

>>>>

\begin{abstract}
Pushdown analysis is better than finite-state analysis in precision and
performance. Why then have we not seen total widespread adoption of these
techniques? For one, the known techniques are technically burdened and
difficult to understand or extend. Control structure of the programming language
gets pulled into the model of computation, which makes extensions to
non-pushdown control structures, such as \rackett{call/cc} or \rackett{shift} and
\rackett{reset}, non-trivial.

We show a derivational approach to abstract interpretation that yields
transparently sound static analyses that can precisely match calls and
returns when applied to well-known abstract machines. We show that
adding memoization and segmenting the continuation into bounded pieces
leads to machines that abstract to static analyses for context-free
reachability by simply bounding the stores. This technique allows us
to derive existing, more technically involved analyses, and a novel
pushdown analysis for delimited, composable control.
\end{abstract}

@@

Once we finitize the address space of the store, this computes a
characteristic finite state machine that has only reachable states of
the pushdown system the unbounded stack model embodies.  To turn this
semantics into PDCFA's algorithm for constructing a Dyck state graph,
we apply a widening operator ${\mathcal F}$ to make $\mstore$,
$\mkont$, and $\mmemo$ shared amongst all states.

@@

\paragraph{Example:} Let's consider a monovariant allocation strategy
(the address for each binding is the variable name itself) to run the
following example:
\begin{lstlisting}[mathescape]
  (let* ([id ($\lambda$ (x) x)]
         [y (id 0)]
         [z (id 1)])
    ($\le$ y z))
\end{lstlisting}
We should expect that a pushdown analysis would predict this evaluates
to true, and there are no loops in the program. 0CFA claims there is a
loop from the second call of \texttt{id} to the first, and thusly
predicts this program evaluates to true or false. PDCFA claims there
is no loop, and depending on the implementation, that the result is
true or false (paper), or just true (implemented)\footnote{This is
  because the paper steps every seen state with the current store
  every iteration, but the implementation only steps states that need stepping.}. Let's take a look at the evaluation after binding
\texttt{id} ($\mstore_0 = [\texttt{id} \mapsto
\set{\vclo{\slam{\texttt{x}}{\texttt{x}}}{\bot}}]$):

@@

@@

CFA2 uses a ``local semantics'' that is similar to our segmenting
continuations at function boundaries, but gets stuck when it gets to a
point where it would need to ``return.'' It instead appeals to an
external, imperative algorithm for summarization to sew the function
calls and returns together by pattern matching on the states that the
local semantics produces. What we show here is almost wholly the same
in character, only embodied still in terms of a machine semantics and
thus more easily reasoned about, and can be run in the concrete.

@@

The way we have been splitting continuations at function calls has
similarities to the meta-continuation approach to modeling delimited
control, given in figure \ref{fig:shift-reset} (adapted from
~\citep{ianjohnson:Biernacki2006274}). We could view each function
call as inserting a prompt, and returns as aborting to the nearest
prompt. Resets then insert a second-tier prompt. The changed semantic
spaces for the shift/reset semantics are as follows:
